Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-11 AnalyticStructure_main.tex
update

The analysis is based on the spectral decomposition of the Liouville equation \eqref{LiouvillianRhopomegaDef1} and the make usage of the \emph{excitation operators}, defined through the 
left and right action eigenvalue equations 
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$. 
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab}
\ee
Excitations are transverse operators so that $\op E_a = \dmnot\op E_a\op Q_0 + \op Q_0\op E_a\dmnot$ and can be expressed through a dyadic representation as follows
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;,
\end{align}
where we have defined the \emph{excited states} $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ that, by construction, are orthogonal to the occupied orbitals of $\hnot$. We observe that
right eigenvalues of $\Liouv$ can be written in terms of the left ones due to the properties of the Liouvillian described in appendix \ref{LiouvillianAction}. 
The introduction of the excitation operators allows us to express the spectral decomposition of the resolvent of the Liouvillian so that:
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
\frac{\trace{\op E_a^t\cdot}}{\omega-\Omega_a} \;.
\ee 
Equations \eqref{ExcitationOperatorsDef1} can be recasted in the form of a Sternheimer-like equations for the excited states, that is   
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\left[-\Omega_a-(\hnot - \eps_p)\right]\ket{\chi_p^a} = \op Q_0\op V'[\op E^t_a]\ket{\psi_p}  \;, 
\end{align}
and the orthonormalization condition \eqref{orthoExcitatioOpDef1} implies %among the excitation operators is expressed as 
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
%The analysis presented in this section aims to clarify the meaning of the formal sum over excitation expressed in \eqref{RhopExcitationDef1} (discrete sum? integral over a continuum?).   
% We restrict our analysis assuming that a set of hypothesis commonly realized in DFT holds (symmetry of the transition basis, specify better? see appendix?). Equations 
% \eqref{ExcitationOperatorsDef1} can be recasted in the form of a Sternheimer-like equations for the excited states, that is   
% \begin{align}\lb{ExcitationOperatorsDef3}
% &\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
% &\left[-\Omega_a-(\hnot - \eps_p)\right]\ket{\chi_p^a} = \op Q_0\op V'[\op E^t_a]\ket{\psi_p}  \;, 
% \end{align}
% and the orthonormalization condition \eqref{orthoExcitatioOpDef1} implies %among the excitation operators is expressed as 
% \be\lb{ExcitedStateOrthNormDef1}
% \sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
% \ee
Once again we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
&\ket{\phi^a_p} = \op G_H(\Omega_a+\eps_p)\ket{s_{\phi}} \;, \nn \\
&\ket{\chi^a_p} = \op G_H(-\Omega_a+\eps_p)\ket{s_{\chi}} \;,
\end{align}
where $\op G_H$ is the Helmholtz Green function, while the (self-consistent) source term $s_\phi$ and $s_{\chi}$ collect the sum of the r.h.s of \eqref{ExcitationOperatorsDef3} 
plus the action of the KS potential on the excited states.

It is possible to show that excitation eigenvalues come in pair and we focus our analysis on solutions with positive energy, \emph{i.e.} with $\Omega_a>0$. Looking at the coordinate representation 
of $\op G_H$ (see eq. \eqref{HelmholtzKernelDef1}) we observe that the Helmholtz kernel evaluated in $-\Omega_a+\eps_p$ contains an exponential damping factor for \emph{all} the positive values of 
$\Omega_a$. As a consequence all the $\ket{\chi^a_p}$ states exhibit a bound-state behavior in the long range. Instead, the asymptotic character of $\op G_H(\Omega_a+\eps_p)$ depends from $\Omega_a$ 
and we can identify the threshold value $\Omega_a=|\eps_p|$ in terms of which $\ket{\phi^a_p}$ changes from a bound-state like behavior to a delocalized spherical-wave like character.  
Furthermore, the orthonormalization conditions \eqref{ExcitedStateOrthNormDef1} act as a ulterior constraint for the bound-state solutions and have the effect of determining the quantization of 
the corresponding energy levels. We conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute a genuine \emph{discrete} set, while 
the ones in the unbound energy range behaves as generalized eigenvectors associated to a continuum spectrum.

@@

% This analysis reveals that the excitations spectrum is composed by two distinct sectors. The first one is realized by the (finite) set of localized excitations with discrete eigenvalues. Extending 
% the results achieved in \cite{boffi2016} for the KS virtual orbitals, to the equations for excited states \eqref{ExcitationOperatorsDef3}, we conclude that localized excitations behave as 
% \emph{observable} quantities and the associated energies possess an absolute meaning, which should not depend from the computational setup used for their representation. 
% 
% Instead, the second sector contains the dense set of delocalized excitations with an associated continuum spectrum. These excitations behave as the unbound virtual orbitals analyzed in \cite{boffi2016} 
% and are deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and 
% the associated spectrum exhibit an explicit dependence from the size of the domain. Excitations belonging to this sector thus lose their intrinsic physical meaning and have to be understood as a tool to 
% express the spectral decomposition of the Liouvillian superoperator.

@@

Susceptibility written as
\be\lb{RhopExcitationDef2}
\sop{{\cal \chi}}(\omega) \cdot   = \left(\sum_a + \int\dd a   \right) \op W_{a} \,
\frac{2\trace{\op W_a \cdot}}{\omega^2-\Omega_a^2} \;,
\ee 
with $\op W_a = \frac{1}{\sqrt{2}}(\op E_a + \op E_a^t)$. 
We have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef2}. 
The elements of the first sector 
correspond to \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

% \subsubsection*{Results of this section}
% 
% From the theoretical analysis of the previous section we can made two statements:
% \begin{itemize}
%  \item Fluctuation states, which constitute the building blocks of the response density, have a bound state-like character up to $\omega=|\eps_h|$ and behave as unbound delocalized
%  wave functions for higher values of $\omega$. 
%  \item The excitation landscape is composed of two sectors with distinct properties. The elements of the first one behave as localized, observable quantities and contribute as discrete 
%  poles to the spectral decomposition of the susceptibility functional. The second sector contains a continuum of delocalized excitations which represents a tool to provide
%  an effective representation of the spectral function. 
% \end{itemize}

% The introduction of the excitation operators allows us to express the susceptibility functional \eqref{SusceptibilityDef1} through the spectral decomposition of the resolvent of the 
% Liouvillian. Pursuing this procedure provides:
% \be\lb{RhopExcitationDef1}
% \sop{{\cal R}}(\omega) \cdot   =
% \left(\sum_E + \int\dd E   \right) \op E \,
% \frac{\trace{\commutator{\dmnot}{\op{\tilde E}}\cdot}}{\omega-E} \;,
% \ee 
% where we have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef1}. 
% The elements of the first sector 
% correspond to \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

@@

@@

The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The left action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouvnot \op O = \commutator{\hnot}{O} \;,
\ee
and
\be\lb{CouplDef1}
\coupl\op O = \commutator{\op V'[\op O]}{\dmnot}  = \op Q_0\op V'[\op O]\dmnot - \dmnot\op V'[\op O]\op Q_0 \;.
\ee
$\op V'$ expresses the perturbation induced by the density dependence on ground state Hamiltonian and can be conveniently expressed by defining the 
\emph{scalar coupling kernel}
\be
U\left[\op O; \op O'\right] \equiv  \int \dd \r \dd \r' \trace{\op O \frac{\delta \hnot }{\delta \rho_0(\r,\r')}
} \trace{\op O' \ket{\r} \bra{\r'}} \;, \nn
\ee
on the basis of which
\be\lb{VpDef1}
\op V'[\op O]= 
\int \dd \r \dd \r' U\left[\ketbra{\r}{\r'};\op O\right] \ketbra{\r'}{\r}\;.
\ee
In view of building the spectral decomposition we are interested in finding both the left and right eigenstates of $\Liouv$.   
The right action of $\Liouv$ can be assessed through the equivalence $\trace{\op O'(\Liouv\op O)} = \trace{(\op O'\Liouv)\op O}$, so that:
\be
\op O\Liouv = -\commutator{\hnot}{\op O} + \int \dd\r\dd\r'
U\left[\commutator{\dmnot}{\op O};\ketbra{\r'}{\r}\right]\ketbra{\r}{\r'} \;. \nn
\ee
In the cases relevant for the present analysis the coupling kernel is symmetric with respect to the exchange of its arguments so also the right action of the 
Liouvillian is expressed in terms of $\op V'$, as follows
\begin{align}\lb{LiouvillianRightActionDef1}
\op O\Liouv &= -\commutator{\hnot}{\op O} + V'\left[\commutator{{\dmnot}}{\op O}\right] = -\commutator{\hnot}{\op O} + \nn \\ 
&+ \dmnot\op V'\left[\commutator{{\dmnot}}{\op O}\right]\op Q_0 + \op Q_0\op V'\left[\commutator{\dmnot}{\op O}\right]\dmnot \;.
\end{align}
In the last equivalence we have made usage of the transverse property (see equation \eqref{RhopTransverseDef1}) of the coupling superoperator. Formula 
\eqref{LiouZeroDef1},\eqref{CouplDef1} and \eqref{LiouvillianRightActionDef1} are the starting point for the solution of the eigenvalues equations \eqref{ExcitationOperatorsDef1}.

>>>>

The analysis is based on the spectral decomposition of the Liouville equation \eqref{LiouvillianRhopomegaDef1} and the make usage of the \emph{excitation operators}, defined through the 
left and right action eigenvalue equations 
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$. 
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations are transverse operators so that $\op E_a = \dmnot\op E_a\op Q_0 + \op Q_0\op E_a\dmnot$ and can be expressed through a dyadic representation as follows
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;,
\end{align}
where we have defined the \emph{excited states} $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ that, by construction, are orthogonal to the occupied orbitals of $\hnot$. 
% We observe that right eigenvalues of $\Liouv$ can be written in terms of the left ones due to the properties of the Liouvillian described in appendix \ref{LiouvillianAction}. 
% The introduction of the excitation operators allows us to express the spectral decomposition of the resolvent of the Liouvillian so that:
% \be\lb{RhopExcitationDef1}
% \sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
% \frac{\trace{\op E_a^t\cdot}}{\omega-\Omega_a} \;.
% \ee 
By applying equations \eqref{LiouZeroDef1},\eqref{CouplDef1} and \eqref{LiouvillianRightActionDef1} presented in appendix \ref{LiouvillianAction} to the dyadic representations of the excitation 
operators, it emerges that equations \eqref{ExcitationOperatorsDef1} can be recasted in the form of a Sternheimer-like equations for the excited states 
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
&\left[-\Omega_a-(\hnot - \eps_p)\right]\ket{\chi_p^a} = \op Q_0\op V'[\op E^t_a]\ket{\psi_p}  \;, 
\end{align}
and the orthonormalization condition \eqref{orthoExcitatioOpDef1} implies %among the excitation operators is expressed as 
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
%The analysis presented in this section aims to clarify the meaning of the formal sum over excitation expressed in \eqref{RhopExcitationDef1} (discrete sum? integral over a continuum?).   
% We restrict our analysis assuming that a set of hypothesis commonly realized in DFT holds (symmetry of the transition basis, specify better? see appendix?). Equations 
% \eqref{ExcitationOperatorsDef1} can be recasted in the form of a Sternheimer-like equations for the excited states, that is   
% \begin{align}\lb{ExcitationOperatorsDef3}
% &\left[\Omega_a-(\hnot - \eps_p)\right] \ket{\phi_p^a} = \op Q_0 \op V'[\op E_a] \ket{\psi_p} \;, \nn\\
% &\left[-\Omega_a-(\hnot - \eps_p)\right]\ket{\chi_p^a} = \op Q_0\op V'[\op E^t_a]\ket{\psi_p}  \;, 
% \end{align}
% and the orthonormalization condition \eqref{orthoExcitatioOpDef1} implies %among the excitation operators is expressed as 
% \be\lb{ExcitedStateOrthNormDef1}
% \sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
% \ee
Once again we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
&\ket{\phi^a_p} = \op G_H(\Omega_a+\eps_p)\ket{s_{\phi}} \;, \nn \\
&\ket{\chi^a_p} = \op G_H(-\Omega_a+\eps_p)\ket{s_{\chi}} \;,
\end{align}
% where $\op G_H$ is the Helmholtz Green function, while the (self-consistent) source term $s_\phi$ and $s_{\chi}$ collect the sum of the r.h.s of \eqref{ExcitationOperatorsDef3} 
% plus the action of the KS potential on the excited states. 
where the source terms $s_\phi$ and $s_{\chi}$ collect the sum of the r.h.s of \eqref{ExcitationOperatorsDef3} plus the action of the KS potential on the excited states.

% The excitation spectrum is symmetric w.r.t the inversion of the eigenvalues $\Omega_a \rightarrow -\Omega_a$ and we focus our analysis on the sector with positive energy, \emph{i.e.} $\Omega_a>0$. 
% Looking at the coordinate representation 
% of $\op G_H$ (see eq. \eqref{HelmholtzKernelDef1}) we observe that the Helmholtz kernel evaluated in $-\Omega_a+\eps_p$ contains an exponential damping factor for \emph{all} the positive values of 
% $\Omega_a$. As a consequence all the $\ket{\chi^a_p}$ states exhibit a bound-state behavior in the long range. Instead, the asymptotic character of $\op G_H(\Omega_a+\eps_p)$ depends from $\Omega_a$ 
% and we can identify the threshold value $\Omega_a=|\eps_p|$ in terms of which $\ket{\phi^a_p}$ changes from a bound-state like behavior to a delocalized spherical-wave like character.  
% Furthermore, the orthonormalization conditions \eqref{ExcitedStateOrthNormDef1} act as a ulterior constraint for the bound-state solutions and have the effect of determining the quantization of 
% the corresponding energy levels. We conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute a genuine \emph{discrete} set, while 
% the ones in the unbound energy range behaves as generalized eigenvectors associated to a continuum spectrum.

The excitation spectrum is symmetric w.r.t the inversion of the eigenvalues $\Omega_a \rightarrow -\Omega_a$ and we focus our analysis on the sector with positive energies. The localization 
character of $\op G_H(\Omega_a+\eps_p)$ depends from $\Omega_a$ and we identify the threshold level $\Omega_a=|\eps_p|$. At the same time we point out that the Helmholtz kernel evaluated in 
$-\Omega_a+\eps_p$ contains an exponential damping factor for \emph{all} the positive values of $\Omega_a$.  As a consequence of these behaviors and assuming the spatial localization of 
the source terms,  we observe that the asymptotic character of $\ket{\phi^a_p}$ changes from bound to unbound as the value of the excitation energy crosses the threshold $|\eps_p|$ while all 
the $\ket{\chi^a_p}$ states exhibit a bound-state behavior for all the values of $a$. 
Furthermore, excited states are subjected to the orthonormalization condition \eqref{ExcitedStateOrthNormDef1} that acts as a ulterior constraint and has the effect of determining the quantization 
of the energy levels associated to the the bound-state solutions. Thus we can conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute 
a genuine \emph{discrete} set, while the ones in the unbound energy range behaves as generalized eigenvectors associated to a continuum spectrum.

@@

@@

%The introduction of the excitation operators allows us to express the spectral decomposition of the resolvent of the Liouvillian so that:
%Susceptibility written as
The introduction of the excitation operators allows us to express the susceptibility functional \eqref{SusceptibilityDef1} through the spectral decomposition of the resolvent of the 
Liouvillian. Pursuing this procedure provides:
\be\lb{RhopExcitationDef2}
\sop{{\cal \chi}}(\omega) \cdot   = \left(\sum_a + \int\dd a   \right) \op W_{a} \,
\frac{2\trace{\op W_a \cdot}}{\omega^2-\Omega_a^2} \;,
\ee 
with $\op W_a = \frac{1}{\sqrt{2}}(\op E_a + \op E_a^t)$. 
We have explicitly separated the contributions of the localized and continuous excitations in the spectral decomposition \eqref{RhopExcitationDef2}. 
The elements of the first sector correspond to \emph{true poles} of the susceptibility while the others can be interpreted only in terms of the spectral function.

@@

This analysis has revealed that the excitations spectrum is composed by two distinct sectors. The first one is realized by the (finite) set of localized excitations with discrete eigenvalues. Extending 
the results achieved in \cite{boffi2016} for the KS virtual orbitals, to the equations for excited states \eqref{ExcitationOperatorsDef3}, we conclude that localized excitations behave as 
\emph{observable} quantities and the associated energies possess an absolute meaning, which should not depend from the computational setup used for their representation.

%  \item The excitation landscape is composed of two sectors with distinct properties. The elements of the first one behave as localized, observable quantities and contribute as discrete 
%  poles to the spectral decomposition of the susceptibility functional. The second sector contains a continuum of delocalized excitations which represents a tool to provide
%  an effective representation of the spectral function.

@@

The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The left action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouvnot \op O = \commutator{\hnot}{O} \;,
\ee
and
\be\lb{CouplDef1}
\coupl\op O = \commutator{\op V'[\op O]}{\dmnot}  = \op Q_0\op V'[\op O]\dmnot - \dmnot\op V'[\op O]\op Q_0 \;,
\ee
where $\op V'$ expresses the perturbation induced by the density dependence on ground state Hamiltonian and can be conveniently expressed by defining the 
\emph{scalar coupling kernel}
\be
U\left[\op O; \op O'\right] \equiv  \int \dd \r \dd \r' \trace{\op O \frac{\delta \hnot }{\delta \rho_0(\r,\r')}
} \trace{\op O' \ket{\r} \bra{\r'}} \;, \nn
\ee
on the basis of which
\be\lb{VpDef1}
\op V'[\op O]= 
\int \dd \r \dd \r' U\left[\ketbra{\r}{\r'};\op O\right] \ketbra{\r'}{\r}\;.
\ee
In view of building the spectral decomposition we are interested in finding both the left and right eigenstates of $\Liouv$.   
The right action of $\Liouv$ can be assessed through the equivalence $\trace{\op O'(\Liouv\op O)} = \trace{(\op O'\Liouv)\op O}$, so that:
\be
\op O\Liouv = -\commutator{\hnot}{\op O} + \int \dd\r\dd\r'
U\left[\commutator{\dmnot}{\op O};\ketbra{\r'}{\r}\right]\ketbra{\r}{\r'} \;. \nn
\ee
In the cases relevant for the present analysis the coupling kernel is symmetric with respect to the exchange of its arguments so also the right action of the 
Liouvillian is expressed in terms of $\op V'$, as follows
\begin{align}\lb{LiouvillianRightActionDef1}
\op O\Liouv &= -\commutator{\hnot}{\op O} + V'\left[\commutator{{\dmnot}}{\op O}\right] = -\commutator{\hnot}{\op O} + \nn \\ 
&+ \dmnot\op V'\left[\commutator{{\dmnot}}{\op O}\right]\op Q_0 + \op Q_0\op V'\left[\commutator{\dmnot}{\op O}\right]\dmnot \;.
\end{align}
In the last equivalence we have made usage of the transverse property (see equation \eqref{RhopTransverseDef1}) of the coupling superoperator. Formula 
\eqref{LiouZeroDef1},\eqref{CouplDef1} and \eqref{LiouvillianRightActionDef1} are the starting point for the solution of the eigenvalues equations \eqref{ExcitationOperatorsDef1}.
